A note on observation processes in epidemic models

### 2019-11-27 ms/ms.tex
minor updates

\maketitle

@@

Despite the amount of effort put into developing better process models, 
disease modelers often neglect details of the observation processes associated with
new disease case reports (often referred to as incidence time series).
Many disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or when an individual becomes symptomatic
(e.g., \cite{bhadra2011malaria, king2015avoidable}); 
some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious). 
\bmb{Maybe mention King et al. -- completely ignored observation processes}

@@

Figure 1A compares the deterministic dynamics of two incidence curves,
$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$ for $\mathcal R_0 = 2$. 
A lag in reporting time delays
the timing of the observed epidemic peak and reduces the size of that peak.
As $\mathcal R_0$ increases, the difference in the size of the peaks increases (Figure 1B)
but the difference in the timing of the peaks decreases (Figure 1C).
For example, when $\mathcal R_0 = 5$, such delay in the reporting of new cases can reduce the size of the observed epidemic peak by almost 50\%.

@@

When we try to estimate all 5 parameters, fitting the recovery model to
infection time series underestimates the basic reproduction number and 
gives a slightly low coverage (Figure 2A). Fitting the infection model to recovery
time series slightly overestimates the basic reproduction number but gives 
good coverage. Fitting the correct models to their corresponding time series
gives unbiased estimates and good coverage.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting the
incorrect model results in a clearer bias (in opposite directions)
and a much lower coverage (cf. \cite{elderd2006uncertainty}).

@@

\bmb{Not including process error. Discuss?}
\bmb{Constant reporting delays don't matter, if system is autonomous}

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, epidemic models often make 
simple assumptions about how cases are reported. 
We used a deterministic SIR model to show that
assumptions about observation processes can
affect the estimates of the basic reproduction number and their associated confidence intervals.

@@

Here, we assumed that the underlying transmission process is deterministic; this 
assumes that all error can be explained by observation errors alone. 
We chose to study a deterministic model SIR for computational efficiency; 
we do not recommend using deterministic models for real outbreak analyses.
Ignoring process errors (i.e., stochasticity in the transmission process) can lead to
overly confident results \citep{king2015avoidable}. Using stochastic models may
give better coverage probabilities even  when wrong observation models are used.
Nonetheless, misspecifying the observation model can still affect conclusions from 
stochastic models and introduce bias.

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		A note on observation processes in epidemic models
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,2,*}
Benjamin M. Bolker\textsuperscript{2,3,4}
\\

\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
\\
\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{4} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

*swp2@princeton.edu
\end{flushleft}

@@

Despite the amount of effort put into developing better process models, 
disease modelers often neglect details of the observation processes associated with
new disease case reports (often referred to as incidence time series).
Many disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or when an individual becomes symptomatic
(e.g., \cite{bhadra2011malaria, king2015avoidable}); 
some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

@@

Figure 1A compares the deterministic dynamics of two incidence curves,
$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$ for $\mathcal R_0 = 2$. 
A lag in reporting time delays
the timing of the observed epidemic peak and reduces the size of that peak.
As $\mathcal R_0$ increases, the difference in the size of the peaks increases (Figure 1B)
but the difference in the timing of the peaks decreases (Figure 1C).
For example, when $\mathcal R_0 = 5$, such delay in the reporting of new cases can reduce the size of the observed epidemic peak by almost 50\%.
Note that $i_2(t)$ essentially assumes that the amount of time between when an individual is infected and when an individual reports the infection is exponentially distributed; a fixed delay in reporting time will not change the shape of an epidemic curve.

@@

When we try to estimate all 5 parameters, fitting the recovery model to
infection time series underestimates the basic reproduction number and 
gives a slightly low coverage (Figure 2A). 
Fitting the infection model to recovery
time series slightly overestimates the basic reproduction number but gives 
good coverage. 
We expect fitting incorrect models to give more biased estimates when $\mathcal R_0$ is higher because the differences in the size of observed epidemic peaks become greater.
Fitting the correct model gives unbiased estimates and good coverage.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting
incorrect models results in a clearer bias (in opposite directions)
and a much lower coverage (cf. \cite{elderd2006uncertainty}).

@@

Mathematical modeling of infectious disease outbreaks helps us understand how disease spreads in a population; 
however, epidemic models often make simple assumptions about how cases are reported. 
We used a deterministic SIR model to show that delays in case reports affect the observed shape of an epidemic curve and the estimates of the basic reproduction number.
Even when the basic reproduction number is small (e.g., $\mathcal R_0 = 2$), fitting incorrect models can introduce small bias and give narrow confidence intervals.

@@

Here, we assumed that the underlying transmission process is deterministic; this 
assumes that all error can be explained by observation errors alone. 
We chose to study a deterministic model SIR for computational efficiency; 
we do not recommend using deterministic models for real outbreak analyses.
Ignoring process errors (i.e., stochasticity in the transmission process) can lead to
overly confident results \citep{king2015avoidable}. Using stochastic models may
give better coverage probabilities even when wrong observation models are used.
Nonetheless, misspecifying the observation model may still affect conclusions from 
stochastic models and introduce bias.
